---
title: 'Reproducible Research: Project 1'
author: "Esteban Castillo"
date: "December 16, 2017"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.


## Steps

1. Load the necessary packages to complete the assignment


```{r echo = TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
library(scales)
library(Hmisc)
```

2. Load the activity dataset for the assignment


```{r echo = TRUE, message=FALSE, warning=FALSE}
if(!file.exists('activity.csv')){
    unzip('repdata%2Fdata%2Factivity.zip')
}
activityData <- read.csv('activity.csv')
```

3. Pre-Process the data into a correct format for the analysis

```{r echo = TRUE, message=FALSE, warning=FALSE}
activityData$date<- as.Date(activityData$date)
```

4. Calculate the total number of steps taken per day

```{r echo = TRUE}
stepsByDay <- tapply(activityData$steps, activityData$date, sum, na.rm=TRUE)
stepsByDay
```

5. Make a histogram of the total number of steps taken each day

```{r echo = TRUE}
qplot(stepsByDay, 
      xlab='Total steps per day', 
      ylab='Frequency using binwith 500',
      binwidth=500)
```

6. Calculate and report the mean and median of the total number of steps taken per day

```{r echo = TRUE}
stepsByDayMean <- mean(stepsByDay)
stepsByDayMean
stepsByDayMedian <- median(stepsByDay)
stepsByDayMedian
```

7. What is the average daily activity pattern?


```{r echo = TRUE}
avg<- tapply(activityData$steps, activityData$interval, mean, na.rm=TRUE)
plot(names(avg), avg, xlab="5-min interval", type="l", ylab="Average no. of steps")
maxavg<- max(avg)
maxavg #5-minute interval
maxinterval<- as.numeric(names(avg)[which(avg==max(avg))])
maxinterval # Max Average Value
```

8. Imputing missing values


- Calculate and report the total number of missing values in the dataset

```{r echo = TRUE}
numMissingValues <- length(which(is.na(activityData$steps)))
numMissingValues #Number of missing values
```

- Devise a strategy for filling in all of the missing values in the dataset: Create a new dataset that is equal to the original dataset but with the missing data filled i

```{r echo = TRUE}
activityDataImputed <- activityData
activityDataImputed$steps <- impute(activityData$steps, fun=mean)
```

- Make a histogram of the total number of steps taken each day

```{r echo = TRUE}
stepsByDayImputed <- tapply(activityDataImputed$steps, activityDataImputed$date, sum)
qplot(stepsByDayImputed, xlab='Total steps per day (Imputed)',
      ylab='Frequency using binwith 500', binwidth=500)
```

- Calculate and report the mean and median total number of steps taken per day.

```{r echo = TRUE}
stepsByDayMeanImputed <- mean(stepsByDayImputed)
stepsByDayMeanImputed
stepsByDayMedianImputed <- median(stepsByDayImputed)
stepsByDayMedianImputed

```

9. Are there differences in activity patterns between weekdays and weekends?


- Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.


```{r echo = TRUE}
activityDataImputed$dateType <-  ifelse(as.POSIXlt(activityDataImputed$date)$wday %in% c(0,6), 
                                        'weekend', 'weekday')
```

- Make a panel plot containing a time series plot


```{r echo = TRUE}
averagedActivityDataImputed <- aggregate(steps ~ interval + dateType, 
                                         data=activityDataImputed, mean)
ggplot(averagedActivityDataImputed, aes(interval, steps)) + 
    geom_line() + 
    facet_grid(dateType ~ .) +
    xlab("5-minute interval") + 
    ylab("avarage number of steps")
```


